---
title: "Working with data"
---

```{r}
#| echo: false
#| message: false
#| results: hide
source(file = "setup-files/setup.R")
```

```{python}
#| echo: false
#| message: false
import shutup;shutup.please()
exec(open('setup-files/setup.py').read())
```

::: {.callout-tip}
## Learning outcomes

- Be able to import tabular data
- Perform basic operations on data

:::

## Libraries and functions

::: {.callout-note collapse="true"}
## Click to expand

::: {.panel-tabset group="language"}
## R

### Libraries
### Functions

:::
:::

## Purpose and aim

In this section we're covering the basics of reading in using tabular data.

## Darwin's finches

We'll look at some data that come from an analysis of gene flow across two finch species [@lamichhaney2020].

The data focus on two species, _Geospiza fortis_ and _G. scandens_. The measurements are split by a uniquely timed event: a particularly strong El Niño event in 1983. This event changed the vegetation and food supply of the finches, allowing F1 hybrids of the two species to survive, whereas before 1983 they could not. The measurements are classed as `early` (pre-1983) and `late` (1983 onwards).

## Reading in data

::: {.panel-tabset group="language"}
## R

There are several functions to read data into R, we're going to use one from the 
`readr` package, which is part of the `tidyverse`. As such, we first need to load 
the package into R's memory, by using the `library()` function:

```{r}
#| message: false
library(tidyverse)
```

This command has to be run every time you start a new R session. Typically you 
want to include the `library()` calls at the top of your script, so that a user 
knows which packages need to be installed to run the analysis.

Our data is provided in CSV format (comma separated values). This format is a 
regular text file, where each value (or column of the table) is separated by a 
comma. To read such a file, we use the `read_csv()` function, which needs at least 
one input: the _path_ of the file we want to read. It is also good practice 
to explicitly define how missing data is encoded in the file with the `na` option. 
In our case, missing data are encoded as an empty string (imagine this as an empty 
cell in a spreadsheet).

Here's the command:

```{r}
finches <- read_csv("data/finches.csv")
```

We see a lot of output there, but this is not an error! It's a message that `read_csv()` 
prints to inform us of what type of data it thinks each column of the data set is. 
We'll discuss this in a while.

It's always useful to have a glimpse at the first few rows of your data set, to see how it is structured. We can do that with the `head()` function.

```{r}
head(finches)
```


### The `data.frame` object

A **data.frame** is the basic type of object that stores _tabular_ data. 
The `readr` package reads data in an "extended" version of a data frame that it 
calls **tibble** (`tbl` for short). The details of their differences are not very 
important unless you are a programmer, but _tibbles_ offer some user conveniences 
such as a better printing method. For the rest of the course we'll refer to 
"data frames" and "tibbles" interchangeably.

:::

## Subsetting data

::: {.panel-tabset group="language"}
## R

We can subset the data in our `finches` table by **column** or **row**. The `tidyverse` package has a series of useful functions that allow you to do this.

### Subsetting by column

We can use the `select()` function to select certain columns, for example if we just wanted the `country` and `year` column. The first argument we give to the function is the data set, followed by the name of the columns we want:

```{r}
select(finches, group, wing)
```


### Subsetting by row

Now let's say we wanted to only keep certain observations - which are organised in rows. Here we can use the `filter()` function. For example, if we only wanted the data for the United Kingdom:

```{r}
filter(finches, species == "G. fortis")
```

Here we've taken the `finches` data set and we asked R to give us the rows where `species == "G. fortis"` is `TRUE`. It goes through all the rows, in this case checking the `species` column. If the statement `species == "G. fortis"` is `TRUE`, it returns the row. Otherwise it doesn't.

We could also use a different conditional statement, for example returning all the rows where the `weight` is larger than 18 grammes:

```{r}
filter(finches, weight > 18)
```
:::

## Chaining commands

Sometimes we need to perform many different operations before we have the right data in the correct format that we need. For example, we might want to filter for certain values and then only keep certain columns. We could perform these operations one by one and save the output of each into an object that we then use for the next operation.

But this is not very efficient. So it can be useful to chain certain operations together, performing them one by one.

::: {.panel-tabset group="language"}
## R

In R we can do this with the **pipe**. We'll be using the pipe operator for tidyverse (`%>%`). The pipe always starts with **data**, which it then "pipes through" to a function.

Let's look at an example, recreating the `filter()` operation we did earlier, but this time with a pipe:

```{r}
finches %>% 
  filter(weight > 18)
```

What it's done is taken the `finches` data set and then sent this to the `filter()` function. The function doesn't need the data set specified explicitly, because it knows it is coming from the pipe.

We can combine this with other functions:

```{r}
finches %>% 
  filter(weight > 18) %>% 
  select(species, weight)
```

Here we've performed the filtering, and then selected the `species` and `weight` columns.

:::

Chaining operations can be a very powerful tool, since it allows you to break down a complex operation into smaller steps. This often makes the analysis a lot less daunting!

## Summary

::: {.callout-tip}
#### Key points

- Tabular data are an excellent format for programming languages
- Having variables in columns and observations in rows makes analysis easier
- We can subset data across columns and rows

:::
